---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)

# ------------------------------------------------------------------------------
# set up for plotting
# ------------------------------------------------------------------------------

theme_set(theme_minimal())

# colour palette
highlight_colour <-  "#35469D"

```
For publishing on github: https://liuyanguu.github.io/post/2021/01/06/rmarkdown-to-github-pages/

## intro

- Question: has domination of tennis by players from the wealthiest countries reduced over the course of the open era (1968 onwards)
- Definitions: 
  - How well a country is doing: proportion of entrants to the first round of Grand Slams (both womens and mens single draw)
  - wealthiest countries: 

## Tennis result data
### Read the data from github

key points
- 2 tours in separate repos
- For each tour each years match results are in a separate csv
- objective: create a dataframe of results for from alls years for each tour

```{r}
#' Get WTA or ATP tour match results from the Tennis Abstract github for the specified year (https://github.com/JeffSackmann)
#'
#' @param year an integer. The earliest years where there are data: wta 1920; atp 1968.
#' @param tour a string either 'wta' or 'atp'
#'
#' @return a dataframe of match results 

get_res <- function(year, tour){

  # form file location string based on year and tour
  repo_url <- glue::glue("https://raw.githubusercontent.com/JeffSackmann/tennis_{tour}/master/")
  file_name <- glue::glue("{tour}_matches_{year}.csv")
  file_location <- str_c(repo_url, file_name)
  #browser()

  # read in data from csv on github
  res_year <- read_csv(file_location,
                      
                      # address issue with column type being read differently
                      # in different files
                      col_types = list(winner_seed = col_double(),
                                         loser_seed = col_double(),
                                       draw_size = col_double())) %>% 
    
    # make variable names consistent
    janitor::clean_names() %>%

    # add identifiers
    mutate(tour = tour, 
           year = year)

  return(res_year)
}

# ------------------------------------------------------------------------------
# download results
# ------------------------------------------------------------------------------

start_year <- 1968
end_year <-  2022

# aggregate atp results into a single dataframe
atp_results <- seq(start_year, end_year) %>% 
  map_df(~get_res(., "atp"))

# aggregate wta results into a single dataframe
wta_results <- seq(start_year, end_year) %>% 
  map_df(~get_res(., "wta"))



```
## Tidy the data

Key points:
- observations are matches
- the players identified as values of the following variables `winner_id`, `winner_name`, `loser_id`, `loser_name`.
- Single variables split across two columns (e.g. `winner_age` and `loser_age` actual a single variable `age`) `winner_ioc` and `loser_ioc`
- Move to observations being a single player's match results

```{r}
#' Get observations relating to either match winners or losers from 
#' tennis abstract match raw result data
#'
#' @param raw_results_df a dataframe of tennis abstract match result raw data
#' @param result specifies whether `winner` or `loser`data should be returned
#'
#' @return a dataframe of match data for either winners or losers

get_data_by_result <- function(raw_results_df, result){
  
  res_to_retain <- result
  
  if(res_to_retain == "winner"){
    res_to_retain_suffix <- "w_"
    res_to_drop <- "loser"
    res_to_drop_suffix <- "l_"
  }
  
  else if (res_to_retain == "loser"){
    res_to_retain_suffix <- "l_"
    res_to_drop <- "winner"
    res_to_drop_suffix <- "w_"
  }
  
  # tidy up data for match losers
  raw_results_df %>%

    # drop columns containing match stats of the winner
    select(-c(starts_with(res_to_drop), starts_with(res_to_drop_suffix))) %>%

    # simplify column naming to enable row binding of winners and losers
    rename_with(.fn = ~str_replace( .x, str_c(res_to_retain, "_"), ""),
                .cols = starts_with(res_to_retain)) %>%
    rename_with(.fn = ~str_replace( .x, res_to_retain_suffix, ""),
                .cols = starts_with(res_to_retain_suffix)) %>%

    # add result column and match_id
    mutate(result = res_to_retain,
           match_id = 1:n(),
           match_id_str = str_c(match_id, tour, year, sep = "_"),
           .after = name)
  
}


#' Puts tennis abstract match raw result data in a tidy form
#'
#' @param raw_results_df a dataframe of tennis abstract match result raw data
#'
#' @return a tidy dataframe of tennis abstract match results data

tidy_match_results <- function(raw_results_df){
  
  losers <- get_data_by_result(raw_results_df, "loser")
  winners <- get_data_by_result(raw_results_df, "winner")
  
  results_tidy <- 
    
    # recombine data on winning and losing players
    bind_rows(losers, winners) %>% 
    
    # order for readability
    arrange(year, tour, match_id) %>% 
    
    # convert variable types where needed
    mutate(tourney_date = lubridate::ymd(tourney_date))
  
  return(results_tidy)
}

# ------------------------------------------------------------------------------
# tidy results
# ------------------------------------------------------------------------------

atp_results_tidy <- tidy_match_results(atp_results)
wta_results_tidy <- tidy_match_results(wta_results)

# create single dataframe with results from both tours
all_results_tidy <- bind_rows(
  atp_results_tidy, wta_results_tidy
)

```

### Focus on relevant data

```{r}

results_focused <- all_results_tidy %>% 
  
  # focus on entrants to grand slam tournaments
  filter(tourney_level == "G") %>% 
  
  # focus on variables of interest
  select(year, tourney_name, tour, name, id, ioc) %>% 
  
  # remove duplicates where players have appeared in multiple rounds 
  unique()

results_focused

```

### Check completeness of data

- I had assumed there had always been 128 plays in each singles draw (so a total of 256 players at each Grandslam)
- Looking smaller draw size before 1989. Checking historical records of tournament draw I could see that some did have mens or womens draws with less than 128 entrants.
- Rather than manually check all thesizes of all the smaller draws are correct, I decided to refine the scope of the analysis to look at Grand Slam tournaments from 1990 onwards (where there is no missing data)
- also identified and issue with tournament names being different for the US Open Men's and Women's Singles in 2019. String matching issue

```{r}
# look for missing data in post 1968 results
results_focused %>% 
  count(year, tourney_name) %>% 
  filter(n != 256)

# focus on Grand Slams from 1990 onwards
results_clean <- results_focused %>% 
  filter(year >= 1990) %>% 
  mutate(tourney_name = replace(tourney_name, 
                                tourney_name == "Us Open",
                                "US Open"))

# confirm there is no missing data for Grand Slams 1990 onwards
num_missing <- results_clean %>% 
  count(year, tourney_name) %>% 
  filter(n != 256) %>% 
  nrow()

assertthat::assert_that(num_missing == 0)


```


## GDP data

### Read in, tidy and focus data

**Source:** [Our World in Data](https://ourworldindata.org/grapher/gdp-per-capita-worldbank?tab=chart)

```{r}
gdp_focus <- read_csv("data/gdp-per-capita-worldbank.csv") %>% 
  
  # make variable names consistent
  janitor::clean_names() %>% 
  
  # rename variables to make them easier to work with 
  rename(gdp_per_capita = gdp_per_capita_ppp_constant_2017_international,
         country = entity,
         country_code = code) %>% 
  
  # focus on 1990 onwards in line with Grand Slam entrant data
  filter(year >= 1990)

# remove entities which are not countries (e.g. regions or continents)
# i.e. those which do not have country codes
non_country_entities <- gdp_focus %>% 
  filter(is.na(country_code)) %>% 
  distinct(country) %>% 
  .$country

gdp_focus <- gdp_focus %>% 
  filter(!country %in% non_country_entities) %>% 
  filter(country != "World")
  
  
```
### Check completeness of data

look for explicitly missing data
and implicitly missing data

```{r}
# -----------------------------------------------------------------------------
# explore explicitly missing data
# -----------------------------------------------------------------------------
visdat::vis_miss(gdp_focus) +
  labs(title = "No explicitly missing data")

# -----------------------------------------------------------------------------
# explore implicitly missing data
# -----------------------------------------------------------------------------

# create a dataframe with all combinations of year and country
gdp_focus_exp_grid <-  expand_grid(country = unique(gdp_focus$country),
            year = unique(gdp_focus$year)) %>%
  
  # add in gdp data where available
  left_join(gdp_focus)

# show implicitly missing data
gdp_missing <- gdp_focus_exp_grid %>% 
  filter(is.na(gdp_per_capita))
gdp_missing

visdat::vis_miss(gdp_focus_exp_grid) +
  labs(title = "But there is implicitly missing data")

```
### exploring and imputing missing data

- where data is missing, in most cases it is missing for a number of years starting at 1990 and then followed by complete data up until 2020. Based on visual inspection of the dataframe and plot (few lines with breaks in the middle) below.

```{r}
# look at countries with missing data
gdp_missing %>% 
  count(country)

countries_gdp_missing <- unique(gdp_missing$country)

# look at where the missing data is in each country's time series
gdp_missing %>% 
  arrange(country, year)

p <- gdp_focus_exp_grid %>% 
  
  # only need to plot countries with missing data
  filter(country %in% countries_gdp_missing) %>% 
  
  # create the plot
  ggplot(aes(year, gdp_per_capita, group = country)) +
    geom_line(alpha = 0.8, colour = highlight_colour)

# make interactive with tooltips for data exploration
plotly::ggplotly(p)
  
```

Given the pattern of the missing data get strange behavior trying to use an interpolate

```{r}
library(zoo)

strange_interpolation <- c("Kosovo", "Cayman Islands", "Iceland")

p <- gdp_focus_exp_grid %>% 
  mutate(
    gdp_per_capita = zoo::na.approx(gdp_per_capita),
    highlight = country %in% strange_interpolation) %>% 
  # only need to plot countries with missing data
  filter(country %in% countries_gdp_missing) %>% 
  
  # create the plot
  ggplot(aes(year, gdp_per_capita, 
             group = country, 
             colour = highlight)) +
    geom_line(alpha = 0.8) +
  scale_colour_manual(values = c("#C7C2C4", highlight_colour)) +
  theme(legend.position = "none")

# make interactive with tooltips for data exploration
plotly::ggplotly(p)
```
So use a simple manual approach. First 

```{r}

```



## Join the two datasets
```{r}
results_clean %>% 
  rename(country_code = ioc) %>% 
  left_join(gdp_clean)
```

