---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
```
For publishing on github: https://liuyanguu.github.io/post/2021/01/06/rmarkdown-to-github-pages/

## intro

- Question: has domination of tennis by players from the wealthiest countries reduced over the course of the open era (1968 onwards)
- Definitions: 
  - How well a country is doing: proportion of entrants to the first round of Grand Slams (both womens and mens single draw)
  - wealthiest countries: 

## Tennis result data
### Read the data from github

key points
- 2 tours in separate repos
- For each tour each years match results are in a separate csv
- objective: create a dataframe of results for from alls years for each tour

```{r}
#' Get WTA or ATP tour match results from the Tennis Abstract github for the specified year (https://github.com/JeffSackmann)
#'
#' @param year an integer. The earliest years where there are data: wta 1920; atp 1968.
#' @param tour a string either 'wta' or 'atp'
#'
#' @return a dataframe of match results 

get_res <- function(year, tour){

  # form file location string based on year and tour
  repo_url <- glue::glue("https://raw.githubusercontent.com/JeffSackmann/tennis_{tour}/master/")
  file_name <- glue::glue("{tour}_matches_{year}.csv")
  file_location <- str_c(repo_url, file_name)
  #browser()

  # read in data from csv on github
  res_year <- read_csv(file_location,
                      
                      # address issue with column type being read differently
                      # in different files
                      col_types = list(winner_seed = col_double(),
                                         loser_seed = col_double(),
                                       draw_size = col_double())) %>% 
    
    # make variable names consistent
    janitor::clean_names() %>%

    # add identifiers
    mutate(tour = tour, 
           year = year)

  return(res_year)
}

# ------------------------------------------------------------------------------
# download results
# ------------------------------------------------------------------------------

start_year <- 1968
end_year <-  2022

# aggregate atp results into a single dataframe
atp_results <- seq(start_year, end_year) %>% 
  map_df(~get_res(., "atp"))

# aggregate wta results into a single dataframe
wta_results <- seq(start_year, end_year) %>% 
  map_df(~get_res(., "wta"))



```
## Tidy the data

Key points:
- observations are matches
- the players identified as values of the following variables `winner_id`, `winner_name`, `loser_id`, `loser_name`.
- Single variables split across two columns (e.g. `winner_age` and `loser_age` actual a single variable `age`) `winner_ioc` and `loser_ioc`
- Move to observations being a single player's match results

```{r}
#' Get observations relating to either match winners or losers from 
#' tennis abstract match raw result data
#'
#' @param raw_results_df a dataframe of tennis abstract match result raw data
#' @param result specifies whether `winner` or `loser`data should be returned
#'
#' @return a dataframe of match data for either winners or losers

get_data_by_result <- function(raw_results_df, result){
  
  res_to_retain <- result
  
  if(res_to_retain == "winner"){
    res_to_retain_suffix <- "w_"
    res_to_drop <- "loser"
    res_to_drop_suffix <- "l_"
  }
  
  else if (res_to_retain == "loser"){
    res_to_retain_suffix <- "l_"
    res_to_drop <- "winner"
    res_to_drop_suffix <- "w_"
  }
  
  # tidy up data for match losers
  raw_results_df %>%

    # drop columns containing match stats of the winner
    select(-c(starts_with(res_to_drop), starts_with(res_to_drop_suffix))) %>%

    # simplify column naming to enable row binding of winners and losers
    rename_with(.fn = ~str_replace( .x, str_c(res_to_retain, "_"), ""),
                .cols = starts_with(res_to_retain)) %>%
    rename_with(.fn = ~str_replace( .x, res_to_retain_suffix, ""),
                .cols = starts_with(res_to_retain_suffix)) %>%

    # add result column and match_id
    mutate(result = res_to_retain,
           match_id = 1:n(),
           match_id_str = str_c(match_id, tour, year, sep = "_"),
           .after = name)
  
}


#' Puts tennis abstract match raw result data in a tidy form
#'
#' @param raw_results_df a dataframe of tennis abstract match result raw data
#'
#' @return a tidy dataframe of tennis abstract match results data

tidy_match_results <- function(raw_results_df){
  
  losers <- get_data_by_result(raw_results_df, "loser")
  winners <- get_data_by_result(raw_results_df, "winner")
  
  results_tidy <- 
    
    # recombine data on winning and losing players
    bind_rows(losers, winners) %>% 
    
    # order for readability
    arrange(year, tour, match_id) %>% 
    
    # convert variable types where needed
    mutate(tourney_date = lubridate::ymd(tourney_date))
  
  return(results_tidy)
}

# ------------------------------------------------------------------------------
# tidy results
# ------------------------------------------------------------------------------

atp_results_tidy <- tidy_match_results(atp_results)
wta_results_tidy <- tidy_match_results(wta_results)

# create single dataframe with results from both tours
all_results_tidy <- bind_rows(
  atp_results_tidy, wta_results_tidy
)

```

### Focus on relevant data

```{r}

results_focused <- all_results_tidy %>% 
  
  # focus on entrants to grand slam tournaments
  filter(tourney_level == "G") %>% 
  
  # focus on variables of interest
  select(year, tourney_name, tour, name, id, ioc) %>% 
  
  # remove duplicates where players have appeared in multiple rounds 
  unique()

results_focused

```

### Check completeness of data

- I had assumed there had always been 128 plays in each singles draw (so a total of 256)

```{r}
results_focused %>% 
  count(year, tourney_name) %>% 
  filter(n != 256)
```


## GDP data

### Read in, tidy and focus data

**Source:** [Our World in Data](https://ourworldindata.org/grapher/gdp-world-regions-stacked-area?country=Sub-Sahara+Africa~Latin+America~Middle+East~South+and+South-East+Asia~East+Asia~Western+Offshoots~Eastern+Europe~Western+Europe)

```{r}
gdp_focus <- read_csv("data/gdp-world-regions-stacked-area.csv") %>% 
  
  # make variable names consistent
  janitor::clean_names() %>% 
  
  # column not needed
  select(-x417484_annotations) %>% 
  
  # focus on years of interest
  filter(year >= start_year) %>% 
  
  # make the dataframe easier to read
  rename(country = entity,
         country_code = code)

# remove entities which are not countries (e.g. regions or continents)
# i.e. those which do not have country codes
non_country_entities <- gdp_focus %>% 
  filter(is.na(country_code)) %>% 
  distinct(country) %>% 
  .$country

gdp_focus <- gdp_focus %>% 
  filter(!country %in% non_country_entities) %>% 
  filter(country != "World")
  
gdp_focus
```
### Check completeness of data

look for 

```{r}
# -----------------------------------------------------------------------------
# explore explicitly missing data
# -----------------------------------------------------------------------------
visdat::vis_miss(gdp_focus) +
  labs(title = "No explicitly missing data")

# -----------------------------------------------------------------------------
# explore implicitly missing data
# -----------------------------------------------------------------------------

# create a dataframe with all combinations of year and country
gdp_focus_exp_grid <-  expand_grid(country = unique(gdp_focus$country),
            year = unique(gdp_focus$year)) %>%
  
  # add in gdp data where available
  left_join(gdp_focus)

# show implicitly missing data
gdp_focus_exp_grid %>% 
  filter(is.na(gdp))

visdat::vis_miss(gdp_focus_exp_grid) +
  labs(title = "But there is implicitly missing data")
```

