---
title: "An animated map showing the countries not represented in Grand Slam tennis (1990-2019)"
output: 
  rmdformats::robobook:
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: false
    highlight: tango
---

**By:** Dr. Chris Martin

**Tools used:** R, R Markdown, `ggplot2`, `gganimate`.

**Techniques used:** animation, mapping.

This notebook produces an animated map which features in my data storytelling project: *The Changing Fortunes of the Richest Countries in Grand Slam Tennis*. You can read the full story [on my website](https://chris-martin-information-design.netlify.app/short-data-stories/tennis).

To produce the chart, I needed data on the women's and men's singles entrants for each Grand Slam tournament since 1990. This came from the excellent [Tennis Abstract](https://github.com/JeffSackmann).

## Setting up the notebook

```{r, warning=FALSE, message=FALSE}
library(tidyverse)

# for mapping
library(sf)
library(rmapshaper)

# for animating ggplot2 charts
library(gganimate)
library(transformr)

# for using custom fonts
library(showtext)

# for table formatting
library(knitr)
```

## Creating the animated map

```{r, warning=FALSE, message=FALSE, cache=TRUE}
# -----------------------------------------------------------------------------
# enable custom fonts
# -----------------------------------------------------------------------------

font_add_google("Lato", family = "lato")
showtext_auto()

# -----------------------------------------------------------------------------
# Read in tennis data from csv (produced by data_clean.Rmd)
# -----------------------------------------------------------------------------

gs_first_round_gdp <- read_csv("../data/results_gdp.csv")

# output to check columns are as expected
gs_first_round_gdp %>% 
  head() %>% 
  kable()

# -----------------------------------------------------------------------------
# Read in spatial data for creating a world map
# -----------------------------------------------------------------------------

countries <- read_sf("../data/ne_110m_admin_0_countries/ne_110m_admin_0_countries.shp") %>% 
  rmapshaper::ms_simplify(keep = 0.5) %>% 
  st_transform(crs = "+proj=eqearth")

# -----------------------------------------------------------------------------
# Clean the data
# -----------------------------------------------------------------------------

# create a grid with all combinations of years and countries
# so we know where data is implicitly missing
fake_data <- expand_grid(year = 1990:2020,
                         ISO_A3_EH = unique(countries$ISO_A3_EH))

# join the tennis data with the spatial data
countries_all_years <- countries %>% 
  select(ISO_A3_EH) %>% 
  filter(ISO_A3_EH != 'ATA') %>% 
  inner_join(fake_data) %>% 
  rename(iso = ISO_A3_EH) %>% 
  left_join(gs_first_round_gdp %>% 
              count(year, country, iso)) %>% 
  mutate(n = if_else(is.na(n), 0L, n),
         at_gs = if_else(n > 0, TRUE, FALSE)) 
  
# -----------------------------------------------------------------------------
# Create the map
# -----------------------------------------------------------------------------
plot <- countries_all_years %>%  
  
  ggplot() + 
    geom_sf(aes(fill = at_gs, colour = at_gs,
                group = seq_along(iso)), 
            size = 0) +
  
    scale_fill_manual(values = c("#C94A54", "#DADADA")) +
    scale_colour_manual(values = c("#C94A54", "#DADADA")) +
  
    # simplify plot background
    theme_void() +
  
    # position and style the text which shows the year advancing
    theme(plot.background = element_rect(fill = "#E9E9E9", colour = NA),
          legend.position = "none",
          text = element_text(family = "lato"),
          plot.title = element_text(face = "bold", size = 45, hjust=0.5,
                                    margin = margin(0,0,5,0), colour = "#2C3130"),
          plot.subtitle = element_text(colour = "#374E83", size = 14, hjust=0.5,
                                    margin = margin(0,0,5,0)),
          plot.margin=unit(c(50,30,30,30), 'pt')) +
  
    # prepare plot object for animation
    transition_time(year) +
    ggtitle('{floor(frame_time)}')

# -----------------------------------------------------------------------------
#  Animate the map
# -----------------------------------------------------------------------------

animate(plot,
        fps = 5,  duration = 30,
        height = 1440, width = 2560, units = "px")

anim_save("not_at_gs.gif")
```




